Validation of marker gene expression

### 2025-02-27 analyses/cell-type-consensus/template-notebooks/marker-gene-validation.Rmd
use fread and back to dplyr

```{r, message=FALSE, warning=FALSE}
# read in consensus and gene expression results 
# use vroom and data.table to try and be more memory-efficient
consensus_dt <- consensus_results_files |> 
  purrr::map(vroom::vroom) |> 
  data.table::rbindlist(fill = TRUE)

gene_exp_dt <- gene_exp_files |> 
  purrr::map(vroom::vroom) |> 
  data.table::rbindlist(fill = TRUE)

@@

```{r, eval=!is_cell_line, message=FALSE}
# combine into one data frame that has all marker genes
# one row for each cell/marker gene combination (total rows= total cells x total marker genes)
# includes original consensus annotation, braod cell type group, marker gene id, logcounts, validation group annotation, information regarding expression of marker gene in CellMarker tissues 
consensus_dt <- consensus_dt |> 
  # add in broad cell type group which is used for plotting
  # groups similar cell types together 
  merge(validation_groups_dt, by = "consensus_annotation", all.x = TRUE) |> 
  merge(gene_exp_dt, by = c("barcodes", "library_id"), all.x = TRUE) |> 
  # add marker gene information (associated validation group annotation, gene observed count, percent tissues)
  # account for the same gene being present in multiple cell types 
  merge(markers_dt, by = "ensembl_gene_id", all.x = TRUE, allow.cartesian = TRUE)

# add a column indicating if gene is present or not
consensus_dt[, detected := logcounts > 0]

@@

```{r, eval=!is_cell_line}
# prep for plots 
# get total number of cells per final annotation group 
total_cells_df <- consensus_dt |> 
  dplyr::select(library_id, barcodes, broad_celltype_group) |> 
  unique() |> 
  dplyr::count(broad_celltype_group, name = "total_cells") 
```

>>>>

```{r, message=FALSE, warning=FALSE}
# read in consensus and gene expression results 
# use vroom and data.table to try and be more memory-efficient
consensus_dt <- consensus_results_files |> 
  purrr::map(fread) |> 
  data.table::rbindlist(fill = TRUE, use.names = TRUE)

gene_exp_dt <- gene_exp_files |> 
  purrr::map(fread) |> 
  data.table::rbindlist(fill = TRUE, use.names = TRUE)

@@

```{r, eval=!is_cell_line, message=FALSE}
# combine into one data frame that has all marker genes
# one row for each cell/marker gene combination (total rows= total cells x total marker genes)
# includes original consensus annotation, braod cell type group, marker gene id, logcounts, validation group annotation, information regarding expression of marker gene in CellMarker tissues 
consensus_dt <- consensus_dt |> 
  # add in broad cell type group which is used for plotting
  # groups similar cell types together 
  dplyr::left_join(validation_groups_dt, by = "consensus_annotation") |> 
  dplyr::left_join(gene_exp_dt, by = c("barcodes", "library_id")) |> 
  # add marker gene information (associated validation group annotation, gene observed count, percent tissues)
  # account for the same gene being present in multiple cell types 
  dplyr::left_join(markers_dt, by = "ensembl_gene_id", relationship = "many-to-many") |> 
  dplyr::mutate(detected = logcounts > 0)

@@

```{r, eval=!is_cell_line}
# prep for plots 
# get total number of cells per final annotation group 
total_cells_df <- consensus_dt |> 
  dplyr::select(library_id, barcodes, broad_celltype_group) |> 
  dplyr::distinct() |> 
  dplyr::count(broad_celltype_group, name = "total_cells") 
```
